2024 US Election Prediction

### 2024-11-04 paper/paper.qmd
Paper updated and references
Updated introduction and abstract, fixed cross references, and added references.bib

---
title: "2024 US Election Prediction"
subtitle: "An Analysis of Donald Trump’s Polling Trends and Predictive Outcomes"
author: 
  - Betty Liu
  - Jingchuan Xu
  - Dingshuo Li
thanks: "Code and data are available at: https://github.com/dawsonlll/2024_US_Election_Analysis"
date: Nov 04, 2024
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format:
  pdf:
    toc: true
  html:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

The 2024 U.S. presidential election is one of the most closely watched events in recent political history. Polling provides a glimpse into the dynamics of the race, offering insights that can shape campaign strategies and public opinion. Among the candidates, former President Donald Trump consistently draws significant attention and debate from the public across different political perspectives. Tracking Trump’s support through the poll reveals his current popularity and potential electoral outcomes. 
The paper dives into Trump’s polling data, analyzing data from various pollsters to see how his support has changed over time. The study explores whether there is an upward or downward trend using linear modelling in Trump’s polling percentage, while also accounting for differences in his support reported by various pollsters. By investigating these patterns, the analysis aims to estimate Trump’s likely level of support as election day approaches.

Estimand: The estimand of this study is the expected level of voter support for Donald Trump in the 2024 U.S. presidential election, based on observed patterns in polling data. This estimated support level provides a snapshot of his projected standing among voters with the adjective by pollster-specific effects.

In this study, we apply statistical models, including a linear model of Trump's percentage over time, to quantify changes in his support level. Additionally, we adjust for pollster variations to provide a more accurate reflection of his support trajectory across different sources. Our findings reveal both temporal trends and variability among pollsters, highlighting the complexities of interpreting polling data in a polarized political environment. These insights contribute to a better understanding of Trump’s standing and the potential outcomes in 2024.

@@

All data analysis was conducted using R @citeR, including statistical computing and graphics. And the following R packages were used: tidyverse @tidy, Palmerpenguins @palmerpenguins, ggplot2 @ggplot2, Dplyr @citedplyr, Knitr @citeknitr, modelsummary @modelsummary, arrow @arrow, here @here, rstanarm @rstanarm, sclaes @scales, and splines @splines. Following  @tellingstories we conducted data simulate, test simulated data, data cleanning, test analysis data, EDA, data modelling. The R code in scripts were adapted from @alexander2023telling

@@

The primary predictor variables for understanding trends in the approval ratings-pct are state and pollscore. The state variable captures the geographic location in which each poll was conducted, providing insight into regional differences in candidate support. By analyzing pct across different states, we can uncover essential patterns and regional variations that may influence or reflect broader national trends. As the @tbl-state-mean-pct shows this variable allows us to compare how support for each candidate fluctuates based on location, revealing possible strongholds or areas of low support that could be pivotal in understanding voter behavior. For instance, we may observe higher variability in pct for swing states compared to states with traditionally stable voting patterns, giving context to national versus regional polling outcomes.
```{r}
#| label: tbl-state-mean-pct
#| tbl-cap: "Preview of States Statistics for Trump and Harris"
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: tbl-preview-pollscore
#| tbl-cap: "Preview of Pollscore"
#| echo: false
#| warning: false
#| message: false
combined_data <- bind_rows(trump_analysis_data, harris_analysis_data)

@@

```{r}
#| label: fig-of-date
#| fig-cap: "figure of PCT Change over Time"
#| fig-pos: "H"
#| echo: false
#| warning: false
#| message: false
combined_data <- trump_analysis_data %>%
  mutate(candidate = "Trump") %>%
  bind_rows(harris_analysis_data %>% mutate(candidate = "Harris"))

@@

## Model Summary and Justification
### Linear Model Summary
The linear model is designed to estimate approval ratings (pct) using key predictors as end_date, pollster. By including end_date, we capture temporal trends in public opinion, which is especially important as election day approaches and public sentiment shifts. A spline function on end_date allows for capturing non-linear trends, enabling the model to reflect fluctuations in approval that may not follow a simple linear trajectory. Including pollsters as a categorical variable controls for systematic differences across pollsters, as different pollsters may use different methods or biases. The summary of linear model for both Trump and Harris is showing at [@appendix-c].

### Bayesian Model Summary
For the Bayesian models, they were employed to estimate approval ratings (pct) for Trump and Harris, using predictors end_date, pollster, state, and pollscore. Bayesian modeling offers flexibility by allowing the incorporation of prior information, which helps stabilize estimates, especially with limited data or high variability. By assigning priors to model parameters, we incorporate prior knowledge while letting the data adjust these estimates, resulting in a balanced interpretation of effects. The model used a spline function on end_date to capture non-linear trends in approval ratings over time, adapting to fluctuations closer to the election date. Priors for each predictor were set as normally distributed with moderate variance, allowing data-driven estimates while avoiding overfitting. The more detailed summart of Bayesian model is in [@appendix-c].

@@

By examining these regional variations, the study reveals that the fate of the election could be determined by a few pivotal swing states, where the balance of support can shift dynamically in response to tailored outreach. For campaign strategists and political analysts, this finding underscores the importance of focusing on swing-state sentiment, which not only reflects the impact of national events on a local scale but also serves as a microcosm of the broader electorate’s potential responsiveness. This nuanced understanding of regional variation illustrates that the election outcome may well rest on the success of capturing undecided voters in these key states, whose shifting allegiances could tip the balance in either candidate’s favor.
﻿

@@

## Google Form
https://docs.google.com/forms/d/e/1FAIpQLSedLCpBggwaUAy1xA2el9yrE3M31_YVVGOnsB1m
EVh73uDahw/viewform?usp=sf_link

# Appendix C {#appendixc}
## Detailed Linear Model Data summary 
```{r}
#| label: tbl-Summary-lin-model
#| tbl-cap: "Summary of Linear Models for Harris and Trump"
#| fig-pos: "H"
#| echo: false
#| warning: false
#| message: false

@@

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

>>>>

---
title: "2024 US Election Prediction"
subtitle: "An Analysis of Donald Trump’s Polling Trends and Predictive Outcomes"
author: 
  - Betty Liu
  - Jingchuan Xu
  - Dingshuo Li
thanks: "Code and data are available at: https://github.com/dawsonlll/2024_US_Election_Analysis"
date: Nov 04, 2024
date-format: long
abstract: "This article analyzes polling data from the 2024 U.S. presidential election to track changes in voter support for Kamala Harris and Donald Trump over time. By applying linear and Bayesian models, we examine trends in their popularity while accounting for differences across pollsters. Our results reveal distinct patterns of support for each candidate, with Harris currently in the lead, and fluctuations influenced by both temporal and regional factors. These insights help illuminate the dynamics of voter sentiment in a polarized political climate, providing valuable perspectives on potential outcomes of the upcoming election."
format:
  pdf:
    toc: true
  html:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

The 2024 U.S. presidential election promises to be one of the most significant and closely watched events in recent political history. Polling data provides a critical window into the dynamics of this campaign, offering insights that shape campaign strategies, inform public opinion, and highlight changing voter sentiment. Among the candidates, former President Donald Trump and Vice President Kamala Harris stand out as each representing different political ideologies that continue to polarize the American electorate. Tracking their approval ratings through polls can not only shed light on their current popularity, but also predict potential electoral outcomes over the course of the campaign.

This study examines polling data for Kamala Harris and Donald Trump, sourced from @polls, to observe changes in voter approval ratings over time. Analyzing these trends through linear and Bayesian models, this paper aims to objectively understand how each candidate’s approval ratings have evolved and identify any clear upward or downward trajectories. A key aspect of this analysis is adjusting for differences between pollsters whose methodologies and regional influences affect reported support levels. These adjustments allow us to gain a more nuanced view of each candidate’s standing, providing insights beyond the results of any single poll.

The main object of this analysis is the expected level of support for Trump and Harris as election day approaches, focusing on trends that are influenced by specific pollsters. By applying statistical models, including linear and Bayesian models of Trump and Harris's percentage support, we aim to quantify changes in support while adjusting for pollster biases. This comprehensive approach allows us to assess temporal patterns and capture changes in support across pollsters.

Our results show that there is a wide range of polling results for both candidates, with different patterns in their support trajectories. Trump's poll percentages show a modest but consistent trend, with some pollsters reflecting more stable support than others. Harris's support fluctuates more significantly, influenced by recent events and regional differences in polling. These insights provide a deeper understanding of each candidate's current position and a basis for predicting potential outcomes.

Estimand: The estimand of this study is the expected level of voter support for Kamala Harris in the 2024 U.S. presidential election, based on observed patterns in polling data. This estimated support level provides a snapshot of her projected standing among voters with the adjective by pollster-specific effects.

@@

All data analysis was conducted using R @citeR, including statistical computing and graphics. And the following R packages were used for conduct code in scripts: tidyverse @tidy, Palmerpenguins @palmerpenguins, ggplot2 @ggplot2, Dplyr @citedplyr, Knitr @citeknitr, modelsummary @modelsummary, arrow @arrow, here @here, rstanarm @rstanarm, sclaes @scales, splines @splines,broom @broom, lubridate @date, janitor @jan and testthat @testthat. Following  @tellingstories we conducted data simulate, test simulated data, data cleanning, test analysis data, EDA, data modelling. The R code in scripts were adapted from @alexander2023telling

@@

The primary predictor variables for understanding trends in the approval ratings-pct are state and pollscore. The state variable captures the geographic location in which each poll was conducted, providing insight into regional differences in candidate support. By analyzing pct across different states, we can uncover essential patterns and regional variations that may influence or reflect broader national trends. As the @tbl-state-mean-pct shows this variable allows us to compare how support for each candidate fluctuates based on location, revealing possible strongholds or areas of low support that could be pivotal in understanding voter behavior. For instance, we may observe higher variability in pct for swing states compared to states with traditionally stable voting patterns, giving context to national versus regional polling outcomes.
```{r}
#| label: tbl-state-mean-pct
#| tbl-cap: "States Statistics for Trump and Harris"
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: tbl-preview-pollscore
#| tbl-cap: "Pollscore for Harris and Trump by Different Pollsters"
#| echo: false
#| warning: false
#| message: false
combined_data <- bind_rows(trump_analysis_data, harris_analysis_data)

@@

```{r}
#| label: fig-of-date
#| fig-cap: "PCT Change over Time"
#| fig-pos: "H"
#| echo: false
#| warning: false
#| message: false
combined_data <- trump_analysis_data %>%
  mutate(candidate = "Trump") %>%
  bind_rows(harris_analysis_data %>% mutate(candidate = "Harris"))

@@

## Model Summary and Justification
### Linear Model Summary
The linear model is designed to estimate approval ratings (pct) using key predictors as end_date, pollster. By including end_date, we capture temporal trends in public opinion, which is especially important as election day approaches and public sentiment shifts. A spline function on end_date allows for capturing non-linear trends, enabling the model to reflect fluctuations in approval that may not follow a simple linear trajectory. Including pollsters as a categorical variable controls for systematic differences across pollsters, as different pollsters may use different methods or biases. The detailed summary of linear model for both Trump and Harris is showing at @tbl-Summary-lin-model in [Appendix -@sec-appendix-c]

### Bayesian Model Summary
For the Bayesian models, they were employed to estimate approval ratings (pct) for Trump and Harris, using predictors end_date, pollster, state, and pollscore. Bayesian modeling offers flexibility by allowing the incorporation of prior information, which helps stabilize estimates, especially with limited data or high variability. By assigning priors to model parameters, we incorporate prior knowledge while letting the data adjust these estimates, resulting in a balanced interpretation of effects. The model used a spline function on end_date to capture non-linear trends in approval ratings over time, adapting to fluctuations closer to the election date. Priors for each predictor were set as normally distributed with moderate variance, allowing data-driven estimates while avoiding over fitting. The more detailed summary of Bayesian model is in [Appendix -@sec-appendix-c]

@@

By examining these regional variations, the study reveals that the fate of the election could be determined by a few pivotal swing states, where the balance of support can shift dynamically in response to tailored outreach. For campaign strategists and political analysts, this finding underscores the importance of focusing on swing-state sentiment, which not only reflects the impact of national events on a local scale but also serves as a microcosm of the broader electorate’s potential responsiveness. This nuanced understanding of regional variation illustrates that the election outcome may well rest on the success of capturing undecided voters in these key states, whose shifting allegiances could tip the balance in either candidate’s favor.

@@

## Google Form Link of Survey
https://docs.google.com/forms/d/e/1FAIpQLSedLCpBggwaUAy1xA2el9yrE3M31_YVVGOnsB1m
EVh73uDahw/viewform?usp=sf_link

# Appendix C {#sec-appendix-c}
## Detailed Linear Model Data summary 
```{r}
#| label: tbl-Summary-lin-model
#| tbl-cap: "Summary of Linear Models for Harris and Trump"
#| fig-pos: "H"
#| echo: false
#| warning: false
#| message: false

@@
